Evaluating \Htwo{} and \COtwo{} Network Strategies for the European Energy System

### 2023-12-18 report/submissions/journal/paper/paper.tex
strip unneeded comments and text

The transition to a carbon-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both \COtwo{} and \Htwo{}, also refered to as carbon and hydrogen, is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including of carbon capture, transport, storage, utilization, and sequestration. Hydrogen offers an efficient way to transport and store energy over long distances and timeframes, and can be produced from electrolysis or steam methane reforming (SMR), for example. Carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels as for example needed for aviation in form of synthetic kerosene from Fischer-Tropsch synthesis, for shipping in form of synthetic methanol and for industrial feedstocks through synthetic methane. From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \COtwo{} and \Htwo{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

@@

% [potentially relevant text blobs from old 3.1]
% CU is preferably deployed in the regions where carbon, hydrogen as well as the inputs and by-product outputs (e.g., waste heat) of the synthesis processes have the lowest cost (or highest value).

% Now we turn to how carbon for CS is sourced by the different models. In the \modBase{}, because of the absence of a carbon network, all carbon that is sequestered (200 Mt, the same in each model) needs to be captured in the regions with access to an offshore CS site. The \modBase{} deploys all available low-cost point-source CC in these regions, and adds DAC in the sequestration locations with the lowest-cost electricity. In other words, the \modBase{} uses the atmosphere as a substitute carbon transport pathway, allowing \CO{} from inland point sources to escape into the atmosphere, and extracting it again from the atmosphere with DAC at the offshore locations.

% In the \modCO{}, the carbon network enables the system to capture carbon from low-cost point sources and transport it to the sequestration sites, trading off CC costs with transport costs.

% In summary, the \COtwo{} network enables the transport of carbon from the lowest-cost CC sources to CU and CS carbon sinks, reducing the need for high-cost DAC.
% The creation of an \Htwo{} network enables the cost-effective delivery of hydrogen from regions with a surplus of low-cost renewable energy to high-demand industrial centers with exogenous hydrogen demand and a surplus of low-cost CC potentials, reducing the need for DAC and for high-cost electrolysis at the industrial centers.
% This highlights how both the implementation of a \COtwo{} or an \Htwo{} network unlocks low-cost carbon capture potentials at geographically fixed point sources across the continent.

@@

\begin{figure*}[ht!]
    \centering
    \includegraphics[width=0.98\linewidth]{comparison/single-technologies/figures/90_nodes/balance_map_dedicated.png}
    \caption{Optimal operation, flows and prices of the carbon (top line) and hydrogen (bottom line) sectors for the \carbonmodel{} (left) and the \hydrogenmodel{} (right) in the net zero scenario. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Lines and arrows show the interregional transportation. Carbon sequestration offshore are drawn as full circles. \carbon{} and \hydrogen{} transport goes from low price to high price areas to supply CU and, in the case of \carbonmodel{}, CS.
    }
    \label{fig:balance_map}
\end{figure*}

To find the reason for why the \Htwo{} network achieves lower system costs than the \COtwo{} network, it is crucial to understand the underlying mechanisms and geographical implications of the two transportation systems. Fig.~\ref{fig:balance_map} maps the average production, consumption and flow in the carbon (top row) and hydrogen (bottom row) sectors for the \modCO{} (left) and \modH{} (right). Each map displays circles indicating the production (top half-circle) and consumption (bottom half-circle) for each region, with lines and arrows representing the average import or export volumes and flow directions between regions. The color shading of each region corresponds to the average price of the carrier.

As we have also discussed above, the \modCO{} captures large amounts of carbon from bioenergy and industrial activities that were untapped in the \modBase{}. In particular, the carbon capture takes place in Central and Eastern Europe at relatively low cost, and almost none of the carbon is used locally (see Fig.~\ref{fig:balance_map}, top left). A large part of the captured carbon is transported from Western Europe to the Iberian Peninsula (70~Mt/h), the British Isles (66~Mt/h) and Denmark (25~Mt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part of the captured carbon is transported to methanation plants in Italy and Greece (25~Mt/h), where the synthetic methane produced is fed into the well-developed gas grid, as Italy serves as a major gas import hub for Europe. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. In absence of an \Htwo{} network, the system places most electrolysis in regions with the best renewable resources to supply local CU ($\sim$1100~GW$_\text{el}$), and smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$75~GW$_\text{el}$). The model also emphasizes gas imports into Central Europe. Hydrogen prices vary strongly across the continent, with low price ``valleys'' in regions with good renewable resources such as the Iberian Peninsula or Denmark with around 60~€/MWh, and high price ``peaks'' in Central Europe with around 95~€/MWh (see bottom left).

@@

% TODO: smoother transition, incl. end of 3.2
% TODO: add network investment costs plot to Appendix. Mention investment H2 > CO2. (The investment in the hydrogen network is four times the investment in the carbon network.)
The \modHybrid{} combines the advantages of hydrogen and carbon networks, resulting in a cost reduction of \label{}5.2\% (\label{}41 billion euros) compared to the \modBase{}, which is lower than the cost reductions of the \COgrid{} and \modH{} (see Fig.~\ref{fig:cost_bar}). In particular, production, consumption and transport in the carbon and hydrogen sectors in the \modHybrid{} combine elements of the \carbonmodel{} and \hydrogenmodel{} (see Fig.~\ref{fig:capacity_maps}). While the topology of the \Htwo{} network roughly corresponds to that of the \modH{}, the \COtwo{} network plays more of a complementary role. In this section, we explore the advantages and disadvantages compared to the other models.

@@

%
In our configuration, the model's time horizon spans one year with a 3-hourly temporal resolution and a spatial resolution of 90 regions. Each of the regions consists of a complex subsystem with technologies for supplying, converting, storing and transporting energy. Exogenous assumptions on energy demand and non-abatable emissions are taken from various sources~\cite{piamanzGeoreferencedIndustrialSites2018,muehlenpfordtTimeSeries2019,mantzosJRCIDEES20152018,NationalEmissionsReported2023,EurostatCompleteEnergyBalance,uwekrienDemandlib2023}. The energy demand for electricity, transport, biomass, heat and gas are defined per region and time-step.
Land transport demand is exogenously assumed to be fully electrified, including heavy-duty vehicles.
Demands for kerosene for aviation, methanol for shipping, and naphtha for industry are not spatially resolved and assumed to be constant throughout all time-steps.
Heat demand is regionally subdivided into shares of urban, rural and industrial sites.
We show the sum of all energy demands in Fig.~\ref{fig:total-demand-bar}.
The system exogenously produces 633~Mt \COtwo{} per year from industry, aviation, shipping and agriculture, 153~Mt of which are fossil-based process emissions.
Location for industrial clusters are taken from~\cite{hotmaps_industrial_db}. Energy demand for industries are calculated from~\cite{mantzosJRCIDEES20152018}.

>>>>

The transition to a carbon-neutral European economy is a pressing challenge that demands coordinated action across various energy sectors. While management of both \COtwo{} and \Htwo{}, also referred to as carbon and hydrogen, is considered a critical component of this transition, a gap exists in understanding how new hydrogen infrastructure effectively interacts with carbon management technologies, including of carbon capture, transport, storage, utilization, and sequestration. Hydrogen offers an efficient way to transport and store energy over long distances and timeframes, and can be produced from electrolysis or steam methane reforming (SMR), for example. Carbon can be stored in geological formations to permanently extract it from the atmosphere, a process known as carbon sequestration (CS). Additionally, carbon can be combined with hydrogen to produce carbonaceous fuels as for example needed for aviation in form of synthetic kerosene from Fischer-Tropsch synthesis, for shipping in form of synthetic methanol and for industrial feedstocks through synthetic methane. From hereon, we jointly refer to the production of these carbonaceous fuels as Carbon Utilization (CU). Carbon can be effectively captured from industrial processes and the combustion of biomass, fossil fuels, or synthetic carbonaceous fuels through point-source carbon capture (CC) techniques, or harvested from the atmosphere using direct air capture (DAC). Both \COtwo{} and \Htwo{} networks are likely to play a crucial role in the cost-effective integration of these technologies to enable net-zero economies.

@@

@@

\begin{figure*}[ht!]
    \centering
    \includegraphics[width=0.98\linewidth]{comparison/single-technologies/figures/90_nodes/balance_map_dedicated.png}
    \caption{Optimal production, consumption, flows and prices of the carbon (top line) and hydrogen (bottom line) sectors for the \carbonmodel{} (left) and the \hydrogenmodel{} (right) in the net zero scenario. For each region, upper semicircles show the average production per technology, lower semicircles the consumption, and colors the average marginal prices. Lines and arrows show the interregional transportation. Carbon sequestration offshore are drawn as full circles. \carbon{} and \hydrogen{} transport goes from low price to high price areas to supply CU and, in the case of \carbonmodel{}, CS.
    }
    \label{fig:balance_map}
\end{figure*}

To find the reason for why the \Htwo{} network achieves lower system costs than the \COtwo{} network, it is crucial to understand the underlying mechanisms and geographical implications of the two transportation systems.
As we have also discussed above, the \modCO{} captures large amounts of carbon from bioenergy and industrial activities that were untapped in the \modBase{}. In particular, the carbon capture takes place in Central and Eastern Europe at relatively low cost, and almost none of the carbon is used locally (see Fig.~\ref{fig:balance_map}, top left). A large part of the captured carbon is transported from Western Europe to the Iberian Peninsula (70~Mt/h), the British Isles (66~Mt/h) and Denmark (25~Mt/h), where it is used along with low-cost electrolytic hydrogen to produce Fischer-Tropsch fuels and methanol. A second part of the captured carbon is transported to methanation plants in Italy and Greece (25~Mt/h), where the synthetic methane produced is fed into the well-developed gas grid, as Italy serves as a major gas import hub for Europe. Finally, a third part of captured carbon is transported directly to sites in the North Sea, Baltic Sea and Mediterranean Sea where it is sequestered. In absence of an \Htwo{} network, the system places most electrolysis in regions with the best renewable resources to supply local CU ($\sim$1100~GW$_\text{el}$), and smaller amounts of electrolyzers where hydrogen is needed for industrial processes across Central Europe ($\sim$75~GW$_\text{el}$). The model also emphasizes gas imports into Central Europe. Hydrogen prices vary strongly across the continent, with low price ``valleys'' in regions with good renewable resources such as the Iberian Peninsula or Denmark with around 60~€/MWh, and high price ``peaks'' in Central Europe with around 95~€/MWh (see bottom left).

@@

The \modHybrid{} combines the advantages of hydrogen and carbon networks, resulting in a cost reduction of \label{}5.2\% (\label{}41 billion euros) compared to the \modBase{}, which is lower than the cost reductions of the \COgrid{} and \modH{} (see Fig.~\ref{fig:cost_bar}). In particular, production, consumption and transport in the carbon and hydrogen sectors in the \modHybrid{} combine elements of the \carbonmodel{} and \hydrogenmodel{} (see Fig.~\ref{fig:capacity_maps}). While the topology of the \Htwo{} network (8~bn€ investments) roughly corresponds to that of the \modH{}, the \COtwo{} network plays more of a complementary role (3~bn€ investments, see Fig.~\ref{fig:cost_bar_transmission}). In this section, we explore the advantages and disadvantages compared to the other models.

@@

%
In our configuration, the model's time horizon spans one year with a 3-hourly temporal resolution and a spatial resolution of 90 regions. Each of the regions consists of a complex subsystem with technologies for supplying, converting, storing and transporting energy. Exogenous assumptions on energy demand and non-abatable emissions are taken from various sources~\cite{piamanzGeoreferencedIndustrialSites2018,muehlenpfordtTimeSeries2019,mantzosJRCIDEES20152018,NationalEmissionsReported2023,EurostatCompleteEnergyBalance,uwekrienDemandlib2023} (see Fig.~\ref{fig:total-demand-bar}). The energy demand for electricity, transport, biomass, heat and gas are defined per region and time-step.
Land transport demand is exogenously assumed to be fully electrified, including heavy-duty vehicles.
Demands for kerosene for aviation, methanol for shipping, and naphtha for industry are not spatially resolved and assumed to be constant throughout all time-steps.
Heat demand is regionally subdivided into shares of urban, rural and industrial sites.
The system exogenously produces 633~Mt \COtwo{} per year from industry, aviation, shipping and agriculture, 153~Mt of which are fossil-based process emissions.
Location for industrial clusters are taken from~\cite{hotmaps_industrial_db}. Energy demand for industries are calculated from~\cite{mantzosJRCIDEES20152018}.
